---
title:  "<h4> <a href='https://github.com/VIS-SIG/Wonderful-Wednesdays/tree/master/data/2021/2021-02-10'> Visualization of patterns of missing data </a> by Agustin Calatroni"
output:
   html_document:
      self_containded: TRUE
      code_download: yes
      toc: false 
---

<a href="https://github.com/agstn/WW" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#70B7FD; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a>

```{=html}
<style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>
```
```{=html}
<style type="text/css">
.main-container {
max-width: 1200px;
margin-left: auto;
margin-right: auto;}
</style>
```
```{r setup, include=FALSE}
options(width = 200)
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE, comment = NA, cache = FALSE)
```

```{r packages}
pacman::p_load(tidyverse)
```

```{r import-data}
d1 <- read.csv("https://raw.githubusercontent.com/VIS-SIG/Wonderful-Wednesdays/master/data/2021/2021-02-10/missing_data.csv") %>%
  select(-contains(".bin.")) %>%
  mutate(
    trt = factor(trt, labels = c("A", "P")),
    gender = factor(gender, labels = c("F", "M"))
  ) %>%
  rowwise() %>%
  mutate(na_num = sum(is.na(c_across(pain.0:pain.10)))) %>%
  mutate(group = case_when(
    na_num > 0  & trt == "P" ~ "P/m",
    na_num > 0  & trt == "A" ~ "A/m",
    na_num == 0 & trt == "P" ~ "P",
    na_num == 0 & trt == "A" ~ "A",) %>% 
      factor(levels = c("A","A/m","P/m","P")))%>%
  ungroup()

write.csv(d1, "ch_missing_data.csv")
```

```{r long-data}
d3 <- d1 %>%
  rownames_to_column(var = 'id') %>%
  pivot_longer(cols = starts_with("pain."),
               values_to = 'pain') %>% 
  separate(name, c("name","time"), "\\.")%>%
  mutate(time = factor(time, levels = c(0:10))) %>% 
  drop_na(pain)
```

```{r seriation}
pacman::p_load(seriation)

d1_na <- d1 %>%
  mutate(across(contains("pain."), ~ is.na(.x) %>% as.numeric())) %>%
  select(contains("pain.")) %>%
  data.matrix()

set.seed(123)
o <- seriate(d1_na, method = "BEA", margin = 1, control = list(rep = 100, istart = 1, jstart = 1))
```

```{r data-reorder}
d2 <- d1 %>%
  slice(get_order(o))
```

```{r patterns-missing, include=FALSE}
pacman::p_load(samon)

# 1. t the time-point
# 2. Number On Study 
# 3. Number Observed 
# 4. Number last seen at time t
# 5. Proportion last seen (of number on-study) 
# 6. Proportion last seen (of number observed) 
# 7. Number intermittent missing data 
# 8. proportion intermittent missing (of number on-study)

# TRT:A
samonTabmat1(d2 %>% filter(trt == "A") %>% select(contains("pain."))) %>%
  samonTable1()
# Check
p1_a <- samonDataCheck(d2 %>% filter(trt == "A") %>% select(contains("pain.")) %>% as.data.frame())
p2_a <-   samonTabmat1(d2 %>% filter(trt == "A") %>% select(contains("pain.")))
p3_a <- data.frame(time = 0:10,
                   miss = 150 - p2_a[,3],
                   mon  = c(sum(p1_a$missingPatterns == '___________'),
                            sum(p1_a$missingPatterns == '*__________'),
                            sum(p1_a$missingPatterns == '**_________'),
                            sum(p1_a$missingPatterns == '***________'),
                            sum(p1_a$missingPatterns == '****_______'),
                            sum(p1_a$missingPatterns == '*****______'),
                            sum(p1_a$missingPatterns == '******_____'),
                            sum(p1_a$missingPatterns == '*******____'),
                            sum(p1_a$missingPatterns == '********___'),
                            sum(p1_a$missingPatterns == '*********__'),
                            sum(p1_a$missingPatterns == '**********_')),
                   int = p2_a[,7])

# TRT:P
samonTabmat1(d2 %>% filter(trt == "P") %>% select(contains("pain."))) %>%
  samonTable1()
# Check
p1_p <- samonDataCheck(d2 %>% filter(trt == "P") %>% select(contains("pain.")) %>% as.data.frame())
p2_p <-   samonTabmat1(d2 %>% filter(trt == "P") %>% select(contains("pain.")))
p3_p <- data.frame(time = 0:10,
                   miss = 150 - p2_p[,3],
                   mon  = c(sum(p1_p$missingPatterns == '___________'),
                            sum(p1_p$missingPatterns == '*__________'),
                            sum(p1_p$missingPatterns == '**_________'),
                            sum(p1_p$missingPatterns == '***________'),
                            sum(p1_p$missingPatterns == '****_______'),
                            sum(p1_p$missingPatterns == '*****______'),
                            sum(p1_p$missingPatterns == '******_____'),
                            sum(p1_p$missingPatterns == '*******____'),
                            sum(p1_p$missingPatterns == '********___'),
                            sum(p1_p$missingPatterns == '*********__'),
                            sum(p1_p$missingPatterns == '**********_')),
                   int = p2_p[,7])
```

```{r heatmap-create}
pacman::p_load(ComplexHeatmap)
pacman::p_load(circlize)

t_column_ha <- HeatmapAnnotation(
  bar1 = anno_barplot(cbind(p3_a['miss'], p3_p['miss']),
                      height = unit(2, "cm"),
                      ylim = c(0, 100),
                      axis_param = list(at=c(25, 50, 100)),
                      gp = gpar(fill = c("#A6CEE3","#B2DF8A"), 
                                col  = c("#A6CEE3","#B2DF8A"))),
  bar2 = anno_barplot(cbind(p3_a['mon'], p3_p['mon']),
                      height = unit(2, "cm"),
                      ylim = c(0,10),
                      gp = gpar(fill = c("#A6CEE3","#B2DF8A"), 
                                col  = c("#A6CEE3","#B2DF8A"))),
  annotation_label = c("Missing\nOverall","Monotone"),
  gap = unit(3, "points")
)

b_column_ha <- HeatmapAnnotation(
  foo1 =  d3 %>% 
    group_by(time, group) %>% 
    summarise(m = mean(pain)) %>% 
    ungroup() %>% 
    pivot_wider(names_from  = group,
                values_from = m) %>%  
    select(-time) %>%  
    anno_lines(
      add_points = TRUE,
      gp    = gpar(col = c("#1F78B4","#A6CEE3","#B2DF8A","#33A02C")),
      pt_gp = gpar(col = c("#1F78B4","#A6CEE3","#B2DF8A","#33A02C"))
    ),
  annotation_name_side = "left",
  annotation_label = c("Mean")
)

col_fun <- circlize::colorRamp2(breaks = c(0, 16, 24, 32, 64), 
                                colors = RColorBrewer::brewer.pal(n = 9, name = "PuOr")[c(1,3,5,7,9)],
                                transparency = 0)

ha_list <- Heatmap(d2 %>%
                     select(contains("pain")) %>%
                     data.matrix(),
                   heatmap_legend_param = list(title = 'Pain',
                                               at = c(0, 15, 24, 32, 64),
                                               labels = c("0-Min", "15-Q1", "24-Q2/M",  "32-Q3", "64-Max")),
                   col = col_fun,
                   na_col = "gray25",
                   column_labels = 0:10,
                   column_names_rot = 0,
                   column_gap = unit(2, "mm"),
                   cluster_columns = FALSE,
                   cluster_rows = FALSE,
                   cluster_row_slices = FALSE,
                   row_split = d2$group,
                   row_title = ' ',
                   row_title_gp = gpar(cex = 0.2),
                   top_annotation    = t_column_ha,
                   bottom_annotation = b_column_ha
) +
  Heatmap(d2 %>% pull(group),
          name = "TRT",
          col = c("#1F78B4","#A6CEE3","#B2DF8A","#33A02C"),
          show_heatmap_legend = FALSE,
          heatmap_legend_param  = list(title = "TRT",
                                       labels = c("Active","A w/Miss", "P w/Miss","Placebo")),
          bottom_annotation = HeatmapAnnotation(ggplot = anno_empty(border = FALSE,
                                                                    height = unit(3, "cm")))
  ) +
  Heatmap(d2 %>% pull(bmi),
          name = "BMI",
          col = circlize::colorRamp2(breaks = c(18, 25, 27, 30, 35), 
                                     colors = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[c(1,3,5,7,9)],
                                     transparency = 0),
          heatmap_legend_param = list(title = "\nBMI (kg/m2)",
                                      at = c(18, 25, 27, 30, 35),
                                      labels = c("18-Min", "25-Q1", "27-Q2",  "30-Q3", "35-Max")),
          bottom_annotation = HeatmapAnnotation(summary = anno_summary(gp = gpar(fill = c("#1F78B4","#A6CEE3","#B2DF8A","#33A02C"))))
  ) +
  Heatmap(d2 %>% pull(age),
          name = "Age",
          col = circlize::colorRamp2(breaks = c(19, 37, 49, 62, 88), 
                                     colors = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[c(1,3,5,7,9)],
                                     transparency = 0),
          heatmap_legend_param = list(title = "\nAge (yr)",
                                      at = c(19, 37, 49, 62, 88),
                                      labels = c("19-Min", "37-Q1", "49-Q2",  "620-Q3", "88-Max")),
          bottom_annotation = HeatmapAnnotation(summary = anno_summary(gp = gpar(fill = c("#1F78B4","#A6CEE3","#B2DF8A","#33A02C"))))
  )+
  Heatmap(d2 %>% pull(gender),
          name = "Gender",
          col = c("#FFCCFF","#CCFFFF"),
          show_heatmap_legend = FALSE,
          heatmap_legend_param  = list(title = "Gender",
                                       at = c("F","M"),
                                       labels = c("Female","Male")),
          bottom_annotation = HeatmapAnnotation(summary = anno_summary(
            border = FALSE,
            axis_param = list(side = 'right',
                              at = c(0, 0.25, 0.5, 0.75, 1),
                              labels = c('0','Female','.5','Male','1'))))
  ) 
```

<style>
div.gray {background-color:#f2f2f2; border-radius: 5px; padding: 5px; margin-right: 350px;}
</style>
<div class = "gray">
**Phase III clinical trial on *Psoriasis* comparing Active Treatment vs Placebo for outcome variable Pain** (collected on a visual analog scale [0-100]). The pain outcome variable was reordered and classified into four distinctive groups: Active Treatment without [   ]{style="background-color: #1F78B4"} with [   ]{style="background-color: #A6CEE3"} any missing data and Placebo without [   ]{style="background-color: #33A02C"} with [   ]{style="background-color: #B2DF8A"} any missing data. The longitudinal pain scores are visualized through a color-coded Heatmap where missing values are coded [   ]{style="background-color: #7f7f7f"}. Vertical and horizontal figures were added to adorn different aspects of the missing data. Vertical top annotations display the *overall* and *monotone* missing data patterns, while bottom annotations display the longitudinal *mean* results for each of the aforementioned groups. Right horizontal figures represent the distribution of covariates extended by the bottom summary display.
</div>
<br>
```{r heatmap-draw, fig.width = 14, fig.height = 9, dpi = 600}
draw(ha_list,
     heatmap_legend_side = 'right',
     gap = unit(c(2,6,6,5), 'mm'))

decorate_annotation("ggplot", {
  vp <- current.viewport()$name
  print(
    ggplot(
      data = d2,
      aes(group, fill = group)) +
      geom_bar() +
      geom_text(stat='count', aes(label=..count..), vjust=1.2, size = 2.5) +
      scale_fill_manual(values = c("#1F78B4","#A6CEE3","#B2DF8A","#33A02C")) +
      labs(y = NULL)+
      theme_void() +
      theme(legend.position = "none"),
    vp = vp
  )
})

decorate_heatmap_body("TRT", {grid.text("Active")},          row_slice = 1)
decorate_heatmap_body("TRT", {grid.text("Active\nw/ Miss")}, row_slice = 2)
decorate_heatmap_body("TRT", {grid.text("Placebo\nw/ Miss")},row_slice = 3)
decorate_heatmap_body("TRT", {grid.text("Placebo")},         row_slice = 4)
```

```{r}
sessioninfo::session_info()%>%
  details::details(
    summary = 'Current session info',
    open    = FALSE
  )
```

#### Links

[ComplexHeatmap](https://jokergoo.github.io/ComplexHeatmap-reference/book/ "https://jokergoo.github.io/ComplexHeatmap-reference/book/complexheatmap-cover.jpg")

[Package Reference](https://jokergoo.github.io/ComplexHeatmap/index.html "http://jokergoo.github.io/complexheatmap_logo.svg")

[seriation: Infrastructure for Ordering Objects Using Seriation](https://cran.r-project.org/web/packages/seriation/index.html)

[samon: Sensitivity Analysis for Missing Data](https://cran.r-project.org/web/packages/samon/index.html)
